Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species

### 2023-06-08 manuscript/manuscript_19May2023.qmd
Major revisions for draft
Revised all datasets and created draft manuscript

  Passive acoustic monitoring is increasingly used for assessing populations of marine mammals; however, analysis of large datasets is limited by our ability to easily classify sounds detected. Classification of beaked whale acoustic events, in particular, require evaluation of multiple lines of evidence by expert analysts. Here we present a highly automated approach to acoustic detection and classification using supervised machine learning and open source software methods. Data from four large scale surveys of beaked whales (North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier). Overall correct classification results ranged from 86% for the South Atlantic data to 99% for the North Atlantic. Results for many species could likely be improved with increased sample sizes and consideration of alternative automated detectors. These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for species and geographic regions.

@@

BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin_acoustic_2017]. This hierarchical random forest event classifier relies on an initial call classifier for each call type (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier. The event classifier considers the probability of assignment to species for each call in the call classifier, along with any event-level characteristics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors. While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin_acoustic_2017]. In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses. This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the overall number of detections was different. This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn2021]. For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.

@@

Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n=70), Sowerby's beaked whales (n=5), True's beaked whales (n=65), and Gervais' beaked whales (n=3). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

For the NAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI, Detector Model: sampsize = 4, ntrees = 30,000; BANTER Model: sampsize = 4, trees=300,000) and provided an overall correct classification rate of 100% for all four species (Fig. 1). There were insufficient samples of Gervais' beaked whales to include in the NAtlantic model (n=2). All classification results were greater than the expected error rate (Prior in Fig. 1a). Results from the EC model are presented here; results from the EC_ICI model can be found in supplementary materials (Supplement Fig. 1).

![BANTER classification results from the NAtlantic dataset (EC). Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/natlantic_Figure.png){fig-align="center"}

The proximity plot (Fig.1b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors. For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result. The importance heat map (Fig. 1c) shows that the most important features for predicting each species.

The strength of the classifications by species can be seen in the plot votes (Fig. 1d), which show the istribution of the votes for each of the 300,000 trials. For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice. In a perfect scenario, all 300,000 trees would have voted for the correct species and therefore the classification would be correct AND the strength of these classifications would be maximized. The plot vote for the NAtlantic dataset show that all species

@@

For the SAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (Detector Model: sampsize = 2, ntrees = 30,000; BANTER Model: sampsize = 1, ntrees = 300,000) and provided an overall correct classification rate of 86.9% for all four species. The sample sizes for the BANTER model was kept low (sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's beaked whale (see Supplement Fig. 3 for EC_ICI_ALT), but resulted in the loss of Sowerby's in the final model. Classification scores ranged from a low of 74.2% (Cuvier's beaked whale) to 100% (Sowerby's and True's beaked whale, Fig. 2); all classification results were greater than the expected error rate (Prior in Fig. 2a). Results from the EC_ICI model are presented here; results from the EC model can be found in supplementary materials (Supplement Fig. 2).

![BANTER classification results from the SAtlantic dataset. Confusion matrix provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate). Proximity plot (a) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (b) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/satlantic_ici_Figure.png){fig-align="center"}

For SAtlantic, the proximity plot (Fig. 2b) shows that True's beaked whales are differentiated from the other species based solely on the two most important features. The importance heat map (Fig. 2c) shows that Cuvier's, Gervais' and Sowerby's beaked whales rely on a number of predictors that are lower in importance and there is likely greater distinction between these species than is represented by the proximity plot (Fig. 2b). The plot votes (Fig. 2d) shows a moderate number of trees in the forest were misclassified and these misclassifications included all species in the model.

@@

The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.1% for all four species. Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, Fig. 3). All classification results were greater than the expected error rate (Prior in Fig. 3a). Results from the EC model are presented here; results from EC_ICI model can be found in supplementary materials (Supplement Fig. 4).

![BANTER classification results from the Hawaii dataset. Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/hawaii_ici_Figure.png)

There are clear distinctions in the primary features used to classify three of the four species as shown in the Proximity Plot (Fig. 3b) which shows the classification results based on the two most 'important' predictors. Distinctive clusters based on the two most important features can be found for Longman's, Blainsville's, and Cuvier's beaked whales, and there is overlap in the feature space of the clusters for Cuvier's and Cross Seamount beaked whales. This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (Fig. 3c), resulting in the low misclassification rates for these species (Fig. 3a confusion matrix). The plot votes graph (Fig. 3d) show strong classification results for Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.

@@

The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States ([@simonis_passive_2020]); species included Baird's beaked whale (n = 31), Unidentified Beaked Whale 'BW43' (n = 135), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 935), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42). BANTER classification trials included (1) echolocation clicks (EC) and (2) echolocation clicks and ICI (EC_ICI).

The best model considered the echolocation clicks and the ICI values (Detector Model: sampsize = 3, ntrees = 10,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 91.4% for all six species. Classification scores ranged from a low of 83.3% (Cross Seamount beaked whale) to 96.9% (Hubbs beaked whale, Fig. 4a). All classification results were greater than the expected error rate (Prior in Fig. 4a). Results from the EC_ICI model are presented here; complete results can be found in supplementary materials (Supplement Figs. 5-7).

![BANTER classification results from the EPacific dataset. Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), priors (expected error rate), and . Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](../Figures/epacific_ici_Figure.png){fig-align="center"}

There is considerable overlap in the primary features used to classify the Cuvier's, Cross Seamount, and BW43 beaked whales, as shown in the Proximity Plot (Fig. 4b), which shows the classification results based on the two most 'important' predictors. ICI was among the most important classification variable for all species (Fig. 4c heat map). The plot votes (Fig. 4d) show strong classification results for most events of all species except Cross Seamount beaked whale.

@@

@fig-meaningless is generated using an R chunk.

```{r, test}
#| label: fig-meaningless
#| fig-cap: A meaningless scatterplot
#| fig-width: 5
#| fig-height: 5
#| fig-align: center
#| out-width: 50%
#| echo: false
#plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in @tbl-simple example.

```{r}
#| label: tbl-simple
#| tbl-cap: Caption centered above table
#| echo: true
knitr::kable(head(mtcars)[,1:4])
```

>>>>

  Passive acoustic monitoring is increasingly used for assessing populations of marine mammals; however, analysis of large datasets is limited by our ability to easily classify sounds detected. Classification of beaked whale acoustic events, in particular, require evaluation of multiple lines of evidence by expert analysts. Here we present a highly automated approach to acoustic detection and classification using supervised machine learning and open source software methods. Data from four large scale surveys of beaked whales (North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier). Overall correct classification results ranged from 87% for the South Atlantic data to 98% for the North Atlantic. Results for many species could likely be improved with increased sample sizes and consideration of alternative automated detectors. These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for species and geographic regions.

@@

BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin_acoustic_2017]. This hierarchical random forest event classifier relies on an initial call classifier for each call type (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier. The event classifier considers the probability of assignment to species for each call in the call classifier, along with any event-level characteristics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors. While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin_acoustic_2017]. In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses. This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the overall number of detections was different. This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn_acoustic_2021]. For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.

@@

Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n=120), Sowerby's beaked whales (n=6), True's beaked whales (n=76), and Gervais' beaked whales (n=3). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

For the NAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI, Detector Model: sampsize = 4, ntrees = 30,000; BANTER Model: sampsize = 4, trees=300,000) and provided an overall correct classification rate of 100% for all four species (@fig-natlantic). There were insufficient samples of Gervais' beaked whales to include in the NAtlantic model (n=2). All classification results were greater than the expected error rate (Prior in @fig-natlantic a). Results from the EC model are presented here; results from the EC_ICI model can be found in supplementary materials (Supplement Fig. 1).

The proximity plot (@fig-natlantic b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors. For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result. The importance heat map (@fig-natlantic c) shows that the most important features for predicting each species.

The strength of the classifications by species can be seen in the plot votes (@fig-natlantic d), which show the istribution of the votes for each of the 300,000 trials. For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice. In a perfect scenario, all 300,000 trees would have voted for the correct species and therefore the classification would be correct AND the strength of these classifications would be maximized. The plot vote for the NAtlantic dataset show that all species

@@

For the SAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (Detector Model: sampsize = 2, ntrees = 30,000; BANTER Model: sampsize = 1, ntrees = 300,000) and provided an overall correct classification rate of 86.9% for all four species. The sample sizes for the BANTER model was kept low (sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's beaked whale (see Supplement Fig. 3 for EC_ICI_ALT), but resulted in the loss of Sowerby's in the final model. Classification scores ranged from a low of 74.2% (Cuvier's beaked whale) to 100% (Sowerby's and True's beaked whale, @fig-satlantic); all classification results were greater than the expected error rate (Prior in @fig-natlantic a). Results from the EC_ICI model are presented here; results from the EC model can be found in supplementary materials (Supplement Fig. 2).

For SAtlantic, the proximity plot (@fig-natlantic b) shows that True's beaked whales are differentiated from the other species based solely on the two most important features. The importance heat map (@fig-natlantic c) shows that Cuvier's, Gervais' and Sowerby's beaked whales rely on a number of predictors that are lower in importance and there is likely greater distinction between these species than is represented by the proximity plot (@fig-natlantic b). The plot votes (@fig-natlantic d) shows a moderate number of trees in the forest were misclassified and these misclassifications included all species in the model.

@@

The most accurate model considered the echolocation clicks only (Detector Model: sampsize = 10, ntrees = 5,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.1% for all four species. Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, @fig-hawaii). All classification results were greater than the expected error rate (Prior in Fig. 3a). Results from the EC model are presented here; results from EC_ICI model can be found in supplementary materials (Supplement Fig. 4).

There are clear distinctions in the primary features used to classify three of the four species as shown in the Proximity Plot (@fig-hawaii b) which shows the classification results based on the two most 'important' predictors. Distinctive clusters based on the two most important features can be found for Longman's, Blainsville's, and Cuvier's beaked whales, and there is overlap in the feature space of the clusters for Cuvier's and Cross Seamount beaked whales. This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (@fig-hawaii c), resulting in the low misclassification rates for these species (@fig-hawaii a confusion matrix). The plot votes graph (@fig-hawaii d) show strong classification results for Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.

@@

The EPacific dataset consisted of 25 drifting recording buoys (20 stations plus 5 seamount experiment drifts) deployed off the west coast of the United States [@simonis_passive_2020]; species included Baird's beaked whale (n = 31), Unidentified Beaked Whale 'BW43' (n = 135), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 935), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42). BANTER classification trials included (1) echolocation clicks (EC) and (2) echolocation clicks and ICI (EC_ICI).

The best model considered the echolocation clicks and the ICI values (Detector Model: sampsize = 3, ntrees = 10,000; BANTER model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 91.4% for all six species. Classification scores ranged from a low of 83.3% (Cross Seamount beaked whale) to 96.9% (Hubbs beaked whale, @fig-epacific a). All classification results were greater than the expected error rate (Prior in @fig-epacific a). Results from the EC_ICI model are presented here; complete results can be found in supplementary materials (Supplement Figs. 5-7).

There is considerable overlap in the primary features used to classify the Cuvier's, Cross Seamount, and BW43 beaked whales, as shown in the Proximity Plot (@fig-epacific b), which shows the classification results based on the two most 'important' predictors. ICI was among the most important classification variable for all species (@fig-epacific c heat map). The plot votes (@fig-epacific d) show strong classification results for most events of all species except Cross Seamount beaked whale.

@@

[![BANTER classification results from the NAtlantic dataset (EC). Confusion matrix (a) provides the percent correct classification or each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represent true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/natlantic_Figure.png){#fig-natlantic="" fig-align="center"}](#fig-natlantic)

[![BANTER classification results from the SAtlantic dataset. Confusion matrix provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (a) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (b) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (c) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/satlantic_ici_Figure.png){#fig-satlantic="" fig-align="center"}](#fig-satlantic)

[![BANTER classification results from the Hawaii dataset. Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/hawaii_ici_Figure.png){#fig-hawaii="" fig-align="center"}](#fig-hawaii)

[![BANTER classification results from the EPacific dataset. Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of variable importance for each species; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.](./images/epacific_ici_Figure.png){#fig-epacific="" fig-align="center"}](#fig-epacific)
